### 2022-06-10 paper.tex
minor change

Sequencing-based assays with single-cell resolution have offered new means to understand the differences between the cells making up a sample. Single-cell RNA sequencing (scRNA-seq) techniques have matured at great pace in recent years. Now, methods to study chromatin are rapidly catching up. Of special interest here is DNA methylation, i.e., the methylation of bases in the DNA, because DNA methylation patterns are passed on to daughter cells in cell division. Especially in mammals, such methylation chiefly occurs at cytosine bases that are immediately after followed by guanine bases, so-called CpG sites.

@@

CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to accidental deamination), and the majority of existing CpG sites are methylated. Presence or absence of the methyl group at CpG sites is believed to affect gene regulation, with methylation inhibiting transcription. In differentiation, methylation of DNA is conjectured to be an important mechanism to silence genome regions that are not needed in the target lineage. One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes.

@@

The standard approach to analyse scBS data is based on methodology developed for the analysis of single-cell RNA-Seq data. Therefore, we will start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analyzed, before we discuss scBS data analysis.

The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene. A first goal is then to assigning cell types or states to cells, and to this end, one needs to establish which cells are similar to each other, i.e., find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile. A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm. In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-5}$) to the normalized fractions before taking the logarithm. Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score. However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts. Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components. As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on very many genes. Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score. Hence, the PCA space representation y are suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

>>>>

Sequencing-based assays with single-cell resolution have offered new means to understand the differences between the cells making up a sample. Single-cell RNA sequencing (scRNA-seq) techniques have matured at great pace in recent years, and methods to study epigenetics at single-cell resolution are rapidly catching up. Of special interest here is DNA methylation, i.e., the methylation of bases in the DNA, which is passed on to daughter cells in cell division. Especially in mammals, such methylation chiefly occurs at cytosine bases that are immediately after followed by guanine bases, so-called CpG \mbox{(5'--C--phosphate--G--3')} sites.

@@

CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to accidental deamination), and the majority of existing CpG sites are methylated. Presence or absence of the methyl group at CpG sites is believed to have effect on gene regulation, with methylation inhibiting transcription. Methylating DNA during differentiation is therefore conjectured to be an important mechanism to silence genome regions that are not needed in the target lineage. One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes.

@@

We start by briefly reviewing the standard approach of analyzing single-cell \emph{RNA}-Seq data, and how this approach is typically adapted to the case of single-cell DNA methylation data.

The starting point in scRNA-Seq is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene. As a first step towards assigning cell types or states to cells, one needs to establish which cells are similar to each other, and to this end, find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile. A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm. In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-5}$) to the normalized fractions before taking the logarithm. Now, one could use Euclidean distances of these vectors  of logarithmized fractions as dissimilarity score. However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts. Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components. As Poisson noise is uncorrelated between genes, it will average out in the top principal components. Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score, which can subsequently be used as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data. The PCA space representation is furthermore used for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).
